---
title: "Coursera Regression Models Course Project"
author: "Written by: Javier Ng"
date: "4th August 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary

In this report, we will examine the `mtcars` data set and explore how miles per gallon (MPG) is affected by different variables. 

Essentially, we’re asking the following question: which types of cars: manual or automatic transmission cars goes the furthest, using (mpg) as an indication.

## Instructions

Motor Trend, an automobile trend magazine is interested in exploring the relationship between a set of variables and miles per gallon (MPG) outcome. In this project, we will analyze the mtcars dataset from the 1974 Motor Trend US magazine to answer the following questions:

* Is an automatic or manual transmission better for miles per gallon (MPG)?
* How different is the MPG between automatic and manual transmissions?

Using simple linear regression analysis, we determine that there is a signficant difference between the mean MPG for automatic and manual transmission cars. Manual transmissions achieve a higher value of MPG compared to automatic transmission. This increase is approximately 2.1 MPG when switching from an automatic transmission to a manual one, with the weight, horsepower and discplacement held constant.

Exploratory analysis and visualizations are located in the Appendix to this document.

## Data Codebook

We analyze the ‘mtcars’ data set through Regression Modelling and exploratory analysis to show how automatic (am = 0) and manual (am = 1) transmissions features affect the MPG feature. The dataset “mtcars” is located in the package “reshape2” first introduced in the Reshaping Data Course of the same Data Specialisation Course.

The data was extracted from the 1974 Motor Trend US magazine, which comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973-74 models).

The data set consists of a data frame with 32 observations (nrow) and 11 variables (ncol).

* mpg: Miles per US gallon
* cyl: Number of cylinders
* disp: Displacement (cubic inches)
* hp: Gross horsepower
* drat: Rear axle ratio
* wt: Weight (lb / 1000)
* qsec: 1 / 4 mile time
* vs: V/S
* am: Transmission (0 = automatic, 1 = manual)
* gear: Number of forward gears
* carb: Number of carburetors

## Data Processing and Transformation

Loading of relevant libraries
```{r}
library(ggplot2)
```

Loading of dataset
```{r}
data(mtcars)
head(mtcars)

# Transform certain variables into factors
mtcars$cyl  <- factor(mtcars$cyl)
mtcars$vs   <- factor(mtcars$vs)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
mtcars$am   <- factor(mtcars$am,labels=c("Automatic","Manual"))
str(mtcars)
```

## Exploratory Data Analysis(EDA)

To help us understand the data, we build exploratory plots. In Appendix under Plot 1, it shows that there is a definite impact on MPG by transmission with Automatic transmissions having a lower MPG.

## Regression Analysis

Now that we have identified that automatic is better for MPG, a simple linear model will be used to quantify this difference. A stepwise model is developed to help select significant predictors for the final model. This step method runs ~lm multiple times and compared between these multiple regression models using both forward selection and backward elimination methods by the AIC (Akaike Information Criterion) stepwise algorithm.

```{R}
firstmodel <- lm(mpg ~ ., data = mtcars)
bestmodel <- step(firstmodel, direction = "both")
```

The best model obtained from the above computations consists of the variables, **wt**, **cyl**, **wt** and **am**. Details of the model are depicted below.

```{R}
summary(bestmodel)
```

From the above, we observe that the Adjusted R^2 value is 0.84 which is the maximum obtained considering all combinations of variables. Thus, we can conclude that more than 84% of the variability is explained by the above model.

Next, we use anova to compare against our base model, that only uses am as a predictor variable, and the best model that was found through performing stepwise selection.

```{R}
basemodel <- lm(mpg ~ am, data = mtcars)
anova(basemodel, bestmodel)
```

As p-value is significant, hence we reject the null hypothesis that the variables **wt**, **hp** and **cyl** do not contribute to the accuracy of the model.

## Residual and Diagnostics

we examine residual plots of our regression model and also compute some of the regression diagnostics of our model to uncover outliers in the data set.

```{R}
par(mfrow = c(2,2))
plot(bestmodel)
```

Here are the following observations from the above plots:

* For the Residuals vs. Fitted plot seem to be randomly scattered on the plot and verify the independence condition.
* The Normal Q-Q plot consists of the points which mostly fall on the line indicating that the residuals are normally distributed.
* The Scale-Location plot consists of points scattered in a constant band pattern, indicating constant variance.
* There are some distinct points of interest (outliers or leverage points) in the top right of the plots.

## Conclusion

In conclusion, Manual transmission cars get more miles per gallon against cars with Automatic transmission.

## Appendix

Boxplot of MPG vs. Transmission

```{R}
boxplot(mpg ~ am, data=mtcars, col=(c("yellow","blue")), xlab="Transmission (0 = Automatic, 1 = Manual)", ylab="Miles per Gallon", main="Boxplot of MPG vs. Transmission")
```

Boxplot of Mileage by Cylinder

```{R}
boxplot(mtcars$mpg ~ mtcars$cyl, data=mtcars, outpch = 19, col=(c("red", "yellow", "blue")), ylab="miles per gallon", xlab="number of cylinders", main="Mileage by Cylinder")
```

Scatter plot matrix for mtcars dataset

```{R}
pairs(mpg ~., mtcars)
```